---
title: "hg19 to hg38"
output: html_document
date: "2023-03-11"
---

The goal of this mark down is to create a better mpra merge file. This file will have both hg19 and hg38 SNPs.

```{r}
library(ggvenn)
```

Insert MPRA directory here
```{r}
# mpra.dir <- "/nfs/jray/screens/ALL_MPRAs/Ho_et_al_analysis/Downstream_Analysis/20240310_mpra_analysis/tcell/glm"
# mpra.name <- "20240310_tcell_glm_mpra_merge"
mpra.dir <- "/nfs/jray/screens/ALL_MPRAs/Ho_et_al_analysis/Downstream_Analysis/20240310_mpra_analysis/stim_jurkat/glm"
mpra.name <- "20240310_stim_jurkat_glm_mpra_merge"
mpra.path <- paste0(mpra.dir,"/data/",mpra.name,"_hg19.txt")
```

Create a bed file for the lift over
```{r}
# Load MPRA data (change data set by pasting different directory)
mpra <- read.table(mpra.path, sep="\t", header=T)
# How many rows in the MPRA data?
nrow(mpra)
```

## 1. Put hg38 names into table

Compare John and Mike's version of the hg38 liftover table
```{r}
#hg38_liftover <- read.table("/nfs/jray/screens/ALL_MPRAs/Ho_et_al_analysis/Downstream_Analysis/hg_19_to_38/data/mpra_snps.hg38_liftover.txt", sep="", header=T)
#hg38_conversion <- read.table("/nfs/jray/screens/ALL_MPRAs/Ho_et_al_analysis/Downstream_Analysis/hg_19_to_38/data/231011_Mouri_hg19_to_hg38_conversion.txt", sep="\t", header=T)
#hg38_liftover$hg19_ld_snp <- hg38_liftover$snpid_hg19
#liftover_merge <- merge(hg38_liftover,hg38_conversion, by="hg19_ld_snp")
#liftover_merge$agree<- liftover_merge$hg38_pos%in%liftover_merge$pos_hg38
#disagree <- subset(liftover_merge, agree==FALSE)
# Differences : 11:118545772:TC:T,17:34839662:C:CG
```

This is the hg19 and hg38 data. We need to put this data into MPRA merge. Make the snpid_hg38 by combining different parts of the exisitng columns
```{r}
hg38_liftover <- read.table("/nfs/jray/screens/ALL_MPRAs/Ho_et_al_analysis/Downstream_Analysis/hg_19_to_38/data/mpra_snps.hg38_liftover.txt", sep="", header=T)
# Get rid of the word "chr" in the chromosome column, so only the number remains
hg38_liftover_chr_mod <- gsub("chr", "", hg38_liftover$chr)
# Get rid of the numbers on the snpid_hg19 to be left with the alleles
hg38_liftover_id_mod <- gsub("[0-9]+", "", hg38_liftover$snpid_hg19)
# Get rid of double colon
hg38_liftover_id_mod <- gsub("::", ":", hg38_liftover_id_mod)
# create the SNP ID out of all the pieces you just created
hg38_id <- paste0(hg38_liftover_chr_mod,":",hg38_liftover$pos_hg38,hg38_liftover_id_mod)
# Make the SNPID into a column
hg38_liftover$hg38_id <- hg38_id
# Give the data frame better column names
colnames(hg38_liftover) <- c("chr38","pos38","SNP19","SNP38")
```

SNP19 and SNP38 must be in the same format for the merge to work


Loading the MPRA
```{r}
# Load MPRA data (change data set by pasting different directory)
mpra <- read.table(mpra.path, sep="\t", header=T)
# How many rows in the MPRA data?
nrow(mpra)
# Making a new column called SNP19 which are the SNPs of genome 19
mpra$SNP19 <- mpra$SNP
```

Merge 38 SNP into mpra
```{r}
# Merge MPRA with the 
new_mpra <- merge(mpra, hg38_liftover, by="SNP19")
# Collection of SNPs in MPRA
set1 <- paste(mpra$SNP)
# Selection of SNPs in the merged MPRA
set2 <- paste(new_mpra$SNP)
# Create a venn diagram of the two SNP list to see which SNPs don't overlap
library(ggvenn)
B <-list("mpra"=set1, "new_mpra"=set2)
names(B) <- c("mpra"," mpra merged with SNP38")
venn_diagram_plot <- ggvenn(B, set_name_size = 5, fill_color = c("springgreen2","forestgreen"),show_elements = TRUE, label_sep = "\n") + 
  ggtitle("mpra vs. mpra merged with SNP38") +
  theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
venn_diagram_plot
```

We now need to change lead_snp, ld_snp, start_snp, end_snp

Replacing columns with columns that say SNP19
```{r}
new_mpra$chr19 <- new_mpra$chr
new_mpra$lead_snp19 <- new_mpra$lead_snp
new_mpra$ld_snp19 <- new_mpra$ld_snp
new_mpra$snp_start19 <- new_mpra$snp_start
new_mpra$snp_end19 <- new_mpra$snp_end
new_mpra$ID19 <- new_mpra$ID

new_mpra <-  subset(new_mpra, select = -c(snp_end,snp_start,lead_snp,ld_snp,ID,chr,SNP,pos,comb))

mpra <- new_mpra
```



This is John's conversion data which has SNP 38 lead_snp, ld_snp
First we need to make sure that this and the previous 
```{r}
hg38_conversion <- read.table("/nfs/jray/screens/ALL_MPRAs/Ho_et_al_analysis/Downstream_Analysis/hg_19_to_38/data/231011_Mouri_hg19_to_hg38_conversion.txt", sep="\t", header=T)
# names(hg38_conversion)
hg38_conversion <- hg38_conversion[,1:11]
# names(hg38_conversion)
names(hg38_conversion) <- c("chr19", "pos19", "rsid", "ref", "alt", "ld_snp19", "lead_snp19","r2", "pos38", "ld_snp38","lead_snp38")
# names(hg38_conversion)
# Manually fixing the ld_snp that is not in John's data but is in Mike's data
for(i in 1:nrow(hg38_conversion)){
 if(hg38_conversion$ld_snp19[i]=="11:118545772:TC:T"){
   hg38_conversion$ld_snp38[i]="11:118675063:TC:T"
 }
   if(hg38_conversion$ld_snp19[i]=="17:34839662:C:CG"){
   hg38_conversion$ld_snp38[i]="17:36483817:C:CG"
 }
}
# test <- subset(hg38_conversion,ld_snp19=="11:118545772:TC:T")

hg38_conversion$SNP19 <- paste0(hg38_conversion$chr19,":",hg38_conversion$pos19,":",hg38_conversion$ref,":",hg38_conversion$alt)

length(unique(hg38_conversion$SNP19))
hg38_conversion <- hg38_conversion[!duplicated(hg38_conversion[,c('SNP19')]),]

names(hg38_conversion) <- c("chr19", "pos19", "rsid_38", "ref_38", "alt_38", "ld_snp19", "lead_snp19","r2_38", "pos38", "ld_snp38","lead_snp38","SNP19")
# names(hg38_conversion)
```


```{r}
hg38_conversion <- subset(hg38_conversion, select = -c(chr19,lead_snp19,ld_snp19,pos38))

# Merge MPRA and conversion table
mpra <- merge(mpra, hg38_conversion, by="SNP19")

mpra <- mpra[!duplicated(mpra$SNP19), ]
# nrow(unique_mpra)

# Now we need to get the names in the new merged correct MPRA correct
# names(mpra)

mpra$r2_19 <- mpra$r2
mpra$rsid_19 <- mpra$rsid
mpra$ID38 <- paste0(mpra$SNP38,":R:wC")
mpra$comb38 <- paste0(mpra$SNP38, "_center_fwd_ref")

# Creating the start snp column by modifying the SNP column. Start snp is calculated by getting the postion then subtracting 1. Then subtracting the absolute value of the difference between the number of reference and alternate alleles.
mpra$snp_start38<-as.numeric(str_split_fixed(mpra$SNP38, "\\:", 4)[,2])-1-abs(nchar(str_split_fixed(mpra$SNP38, "\\:", 4)[,3])-nchar(str_split_fixed(mpra$SNP38, "\\:", 4)[,4]))
# Create a end_snp column which modifies the SNP column and the values after the first colon and before the second colon
mpra$snp_end38<-as.numeric(str_split_fixed(mpra$SNP38, "\\:", 4)[,2])
# names(mpra)

mpra <- subset(mpra, select = -c(r2,rsid))
names(mpra)
```

Write the table with both hg19 and hg38
```{r}
write.table(mpra, paste0(mpra.dir,"/data/",mpra.name,"_hg38_hg19.txt"), row.names=F, col.names=T, sep="\t", quote=F)
```

Creating a permanent hg38 without special column names
```{r}
mpra <- subset(mpra, select = -c(r2_19,rsid_19,SNP19,lead_snp19,ld_snp19,snp_start19,snp_end19,pos19,chr19,ID19,ref_allele,alt_allele))

mpra$r2 <- mpra$r2_38
mpra$rsid <- mpra$rsid_38
mpra$SNP <- mpra$SNP38
mpra$ld_snp <- mpra$ld_snp38
mpra$lead_snp <- mpra$lead_snp38
mpra$pos <- mpra$pos38
mpra$chr <- mpra$chr38
mpra$ref_allele <- mpra$ref_38
mpra$alt_allele <- mpra$alt_38
mpra$ID <- mpra$ID38

mpra$comb <- mpra$comb38
mpra$snp_start <- mpra$snp_start38
mpra$snp_end <- mpra$snp_end38

mpra <- subset(mpra, select = -c(r2_38,rsid_38,SNP38,lead_snp38,ld_snp38,pos38,chr38,ID38,ref_38,alt_38,comb38,snp_start38,snp_end38))

names(mpra)
# May need to add or subtract 2 depending on the amount of columns in the table
mpra <- mpra[,c(74,78,77,79,80,81:84,75,76,1:73)]
# mpra <- mpra[,c(72,76,75,77,78,79:82,73,74,1:71)]
names(mpra)
```

```{r}
write.table(mpra, paste0(mpra.dir,"/data/",mpra.name,"_hg38.txt"), row.names=F, col.names=T, sep="\t", quote=F)

mpra_hg38 <- read.table(paste0(mpra.dir,"/data/",mpra.name,"_hg38.txt"), sep="\t", header=T)
```

#################

END OF NORMAL SCRIPT

#################

Check on the variants that didn't agree between John and Mike's liftover (11:118545772:TC:T,17:34839662:C:CG)
```{r}
mpra38 <- read.table(paste0(mpra.dir,"/data/",mpra.name,"_hg38.txt"), sep="\t", header=T)
disagree_var1 <- subset(mpra38, SNP=="11:118675063:TC:T")
disagree_var1
disagree_var2 <- subset(mpra38, SNP=="17:36483817:C:CG")
disagree_var2
```




