---
title: "MPRA merge creation"
output: html_document
author: Max Dippel
date: "2024-03-11"
---

This is the document I will use to create the mpra merge files for all of the cell types for the March 10th mpra pipeline run.
There will be a place for importing the mpra in which will have to change every time. 

MPRA files ran through this pipeline:
OLJR.C_Tcell_emVAR_glm_20240310.out
OLJR.C_Tcell_emVAR_ttest_20240310.out
OLJR.A_Jurkat_emVAR_glm_20240310.out
OLJR.D_Jurkat.Stim_emVAR_glm_20240310.out

MPRA files which need to be ran through this pipeline:
OLJR.B_GM12878_emVAR_glm_20240310.out

Hash and unhash the lines at 478 and 548 to prevent writing tables you do not want

Import MPRA, inspect data, and set directories
```{r, eval=FALSE}
# Load the .out file I get from Jax which is just the MPRA data
mpra_out<-read.delim("/nfs/jray/screens/ALL_MPRAs/Ho_et_al_analysis/Primary_Data/20240310_analysis/results/OLJR.C_Tcell_emVAR_glm_20240310.out",header=T, stringsAsFactors = F,sep="\t")
# How many rows are in that file?
nrow(mpra_out)
# How many columns are in that file?
ncol(mpra_out)
# What are the column names in the mpra?
names(mpra_out)
# Directory to put the mpra in (with cell type and glm or ttest and no slash at the end)
mpra.dir <- "/nfs/jray/screens/ALL_MPRAs/Ho_et_al_analysis/Downstream_Analysis/20240310_mpra_analysis/tcell/glm"
# Name of MPRA file (end with hg19 for this Rmd)
mpra.name <- "20240310_tcell_glm_mpra_merge_hg19.txt"
# This decides what option to choose for the variant category filter. There are currently two options "mouri" and "tcell
filter <- "tcell"
```

This code will make columns the same name as the analysis I'm about to run so things are not annoying.
```{r, eval=FALSE}
# Change the name of the columns to match the functions
mpra_out$LogSkew <- mpra_out$Log2Skew
# Change the name of the column to match the functions
mpra_out$A.Ctrl.Mean <- mpra_out$A_Ctrl_Mean
mpra_out$A.Exp.Mean <- mpra_out$A_Exp_Mean
mpra_out$A.log2FC <- mpra_out$A_log2FC
mpra_out$A.log2FC_SE <- mpra_out$A_log2FC_SE
mpra_out$A.logP <- mpra_out$A_logP
mpra_out$A.logPadj_BH <- mpra_out$A_logPadj_BH
mpra_out$A.logPadj_BF <- mpra_out$A_logPadj_BF

mpra_out$B.Ctrl.Mean <- mpra_out$B_Ctrl_Mean
mpra_out$B.Exp.Mean <- mpra_out$B_Exp_Mean
mpra_out$B.log2FC <- mpra_out$B_log2FC
mpra_out$B.log2FC_SE <- mpra_out$B_log2FC_SE
mpra_out$B.logP <- mpra_out$B_logP
mpra_out$B.logPadj_BH <- mpra_out$B_logPadj_BH
mpra_out$B.logPadj_BF <- mpra_out$B_logPadj_BF
mpra_out$Skew.logP <- mpra_out$Skew_logP
mpra_out$Skew.logFDR <- mpra_out$Skew_logFDR
# This column creates the project column
mpra_out$project <- "TGWAS"
# This is a new column I am adding for this analysis
mpra_out$Skew.logFDR_act <- mpra_out$Skew_logFDR_act
# adding this also. I used to do it higher in the analysis.
# The  linkage analysis will not run without this. Is this okay?
mpra_out$ld_snp<-mpra_out$SNP
# eliminate all of the columns that have the names we don't want
mpra <- subset(mpra_out, select = -c(Log2Skew,A_Ctrl_Mean,A_Exp_Mean,A_log2FC,A_log2FC_SE,A_logP,A_logPadj_BH, A_logPadj_BF,B_Ctrl_Mean,B_Exp_Mean,B_log2FC,B_log2FC_SE,B_logP,B_logPadj_BH,B_logPadj_BF,Skew_logP,Skew_logFDR,Skew_logFDR_act))
```

# MPRA merge

Load the packages
```{r packages, eval=FALSE, message=FALSE}
library(stringr)
library(IRanges)
library(SummarizedExperiment)
library(data.table)
library(readxl)
library(biomaRt)
```

```{r}
nrow(mpra)
```


```{r ld data, eval=FALSE}
#Merge LD files across chromosomes
# Create an empty data frame
dat.ld.all<-data.frame(lead_snp=character(), ld_snp=character(), r2=numeric(), stringsAsFactors = F)
# A loop that says "for every chromosome..."
for(c in c(1:22, "X")){
  # read in Mike's linkage files for each chromosome
  dat.ld.chr<-read.delim(paste("/nfs/jray/screens/Mouri_et_al_MPRA/mhguo1-T_cell_MPRA-5c36361/annotate_mpra/ld/ld/mpra.chr", c, ".ld", sep=""), header=T, stringsAsFactors = F, sep="")
  # Select the columns you need from this data set
  dat.ld.chr<-subset(dat.ld.chr, select=c(SNP_A, SNP_B, R2))
  # Rename those columns 
  names(dat.ld.chr)<-c("lead_snp","ld_snp", "r2")
  # combine these values with the empty data frame you created earlier
  dat.ld.all<-rbind(dat.ld.all, dat.ld.chr)
  # Remove dat.ld.chr
  rm(dat.ld.chr)
}

#remove duplicates. If proxy SNP has multiple entries, select the one with the highest r2
dat.ld.all<-dat.ld.all[order(dat.ld.all$ld_snp, -dat.ld.all$r2),]
dat.ld.all<-dat.ld.all[!duplicated(dat.ld.all$ld_snp),]

#Merge in LD information
mpra<-merge(mpra, dat.ld.all,  by="ld_snp", all.x=T, all.y=F)

```

```{r}
nrow(mpra)
```

This big loop to integrate the PICS (probabilistics identification of causal snps) data into the mpra.merge data set 
```{r PICS data, eval=FALSE}
# Order of GWAS for plotting
gwas.order<- c("Crohns","MS","Psoriasis", "RA","T1D","UC", "IBD") 
# Merge in GWAS and PICS fine-mapping data
for(d in gwas.order){
  # Read in PICS results
  gwas<-read.delim(paste0("/nfs/jray/screens/Mouri_et_al_MPRA/mhguo1-T_cell_MPRA-5c36361/pics/run/", d, ".ld_0.8.pics"), header=F, stringsAsFactors = F, sep="\t")
  # Select certain columns in the gwas data
  gwas<-gwas[,c(1,2,4,6,7)]
  # Name the columns of the gwas data
  names(gwas)<-c("lead_snp", "ld_snp", "r2", "pics","pval")
  # eliminate the letters "chr" from lead and lp snp
  gwas$lead_snp<-gsub("chr", "", gwas$lead_snp)
  gwas$ld_snp<-gsub("chr", "", gwas$ld_snp)
  
  #Merge PICS results with MPRA to subset for SNPs tested in the MPRA
  gwas<-merge(gwas, mpra[,c("ld_snp", "lead_snp")], by=c("ld_snp", "lead_snp"), all.x=F, all.y=F)

  #Remove SNPs seen multiple times, picking the SNP with the most significant p-value in catalog, then the highest PICS posterior probability
  gwas<-gwas[order(-gwas$pval, -gwas$r2),]
  gwas<-gwas[!duplicated(gwas[,c("ld_snp", "lead_snp")]),]
  
# Generate credible sets
  # Order the data by SNPs
  gwas <- gwas[order(gwas$lead_snp, -gwas$pics),]
  gwas$PP_running<-0 #Create variable with a running sum of PICS probabilities
  gwas$PP_sum <- 0 #Create variable with sum of PICS probablities
  #Note: need this second variable to help break ties (if two SNPs have the same PP, the automatically will have the same credible set annotation)

  prev_var <- "dummy_variable" #Create variable to test whether we've moved onto a new locus
  for(i in c(1:nrow(gwas))){
    # Creating a new variable from lead snp
    new_var<-gwas[i,]$lead_snp
    # if the pics value is greater than zero
    if(gwas[i,]$pics>0){
      if(new_var==prev_var){ #Test if next variant is still in same locus
        gwas[i,]$PP_sum<-gwas[i-1,]$PP_sum+gwas[i,]$pics #Add pics to running PP sum
        
        if(gwas[i,]$pics==gwas[i-1,]$pics){ #If two variants have the same PP, force them to have the same PP_running
          gwas[i,]$PP_running<-gwas[i-1,]$PP_running
        }else{
          gwas[i,]$PP_running<-gwas[i,]$pics+gwas[i-1,]$PP_sum #If not Add pics and PP_sum to running PP sum
        } 
      }else{ #If this variant is in a new locus, assign the PICS PP as their PP_sum and PP_running
        gwas[i,]$PP_sum<-gwas[i,]$pics 
        gwas[i,]$PP_running<-gwas[i,]$pics
      }
    }
    prev_var<-new_var #If lead SNP is different, we've now moved on to new locus
  }
  
# Merge with MPRA data
  # subseet the gwax data to a few variants
  gwas<-subset(gwas, select=c(ld_snp, pval,pics,PP_running))
  # Round the gwas pvalue to two digits
  gwas$pval<-round(gwas$pval,2)
  # Creating a disease specific name with underscores for the pics statistics
  names(gwas)[2:4]<-paste0(d, "_", names(gwas)[2:4])
  # merge he gwas data and mpra by ld_snp
  mpra<-merge(mpra, gwas, by="ld_snp", all.x=T, all.y=F)
  # remove the gwas object
  rm(gwas)
}
```

```{r}
nrow(mpra)
```

Creating bed-file like columns
```{r}
# Make the MPRA file into a bed-like file for downstream annotations
# Creating the start snp column by modifying the SNP column. Start snp is calculated by getting the postion then subtracting 1. Then subtracting the absolute value of the difference between the number of reference and alternate alleles.
mpra$snp_start<-as.numeric(str_split_fixed(mpra$SNP, "\\:", 4)[,2])-1-abs(nchar(str_split_fixed(mpra$SNP, "\\:", 4)[,3])-nchar(str_split_fixed(mpra$SNP, "\\:", 4)[,4]))
# Create a end_snp column which modifies the SNP column and the values after the first colon and before the second colon
mpra$snp_end<-as.numeric(str_split_fixed(mpra$SNP, "\\:", 4)[,2])
# Create a chromosome column by modifying the SNP column. Taking everything before the first colon and adding it to the letter "chr"
mpra$chr<-paste0("chr", str_split_fixed(mpra$SNP, "\\:", 4)[,1])
# Create a strand column
mpra$strand<-"+"
```

```{r}
nrow(mpra)
```

DHS (DNase 1 hypersensitive sites) data 
```{r}
### DHS data
# Eliminating all the columns that have "dhs_" in them which happen to be none of the columns
mpra<-mpra[,!(grepl("dhs_", names(mpra)))]
# Make a granges object from the mpra spreadsheet
mpra.se<-makeGRangesFromDataFrame(mpra,  seqnames = "chr", start.field = "snp_start", end.field = "snp_end",keep.extra.columns=TRUE)

# Read in DHS position file
dhs.pos<- read.delim(gzfile("/nfs/jray/screens/ALL_MPRAs/Ho_et_al_analysis/Downstream_Analysis/2020_mouri_public_data_replication/data/original_hg19_dhs_data/DHS_Index_and_Vocabulary_hg19_WM20190703.txt.gz"), header=T, stringsAsFactors = F)
# Only keep the first three columns of the DHS pos data
dhs.pos<-dhs.pos[,c(1:3)]
# Make peak column which puts together the chromosome, the start and the end
dhs.pos$peak<-paste0(dhs.pos$seqname, ":", dhs.pos$start, "_", dhs.pos$end)
# make a granges from the dhs pos data
dhs.pos.se<-makeGRangesFromDataFrame(dhs.pos, seqnames = "seqname", start.field = "start", end.field = "end",keep.extra.columns=TRUE)

#Find DHS peaks that overlap with MPRA
overlap_peaks<-data.frame(subsetByOverlaps(dhs.pos.se, mpra.se, type="any"), stringsAsFactors = F)$peak
overlap_index<-which(dhs.pos$peak%in%overlap_peaks)

# bring in dhs data
dhs.dat<-fread("/nfs/jray/screens/ALL_MPRAs/Ho_et_al_analysis/Downstream_Analysis/2020_mouri_public_data_replication/data/original_hg19_dhs_data/dat_bin_FDR01_hg19.txt.gz")
#Subset for only DHS peaks that overlap with MPRA in the dhs data
dhs.dat<-dhs.dat[overlap_index,]
#Subset for only DHS peaks that overlap with MPRA in the dhs pos data
dhs.pos<-dhs.pos[overlap_index,]
# Now combin ehte dhs and dhs pos data
dhs.dat<-cbind(dhs.pos[,c(1:3)],dhs.dat)

#Read in sample information
sample.dat<-read.delim("/nfs/jray/screens/ALL_MPRAs/Ho_et_al_analysis/Downstream_Analysis/2020_mouri_public_data_replication/data/original_hg19_dhs_data/DHS_Index_and_Vocabulary_metadata.tsv", sep="\t", header=T, stringsAsFactors = F)
# Select certain columns in the sample data
sample.dat<-subset(sample.dat, !is.na(library.order),select=c(library.order,Biosample.name, System,Subsystem,  Organ, Biosample.type, Biological.state))

#Add peak index into MPRA for faster calculation
# Create three empty columns in mpra
mpra$peak_index1<-NA
mpra$peak_index2<-NA
mpra$peak_index3<-NA
# Create a row number column for the dhs data sets
dhs.pos$index<-seq(1:nrow(dhs.pos))
dhs.dat$index<-seq(1:nrow(dhs.dat))

# These hashed out portions are hashed out in the original code. I am going to leave them there.
# For every row in mpra, create a subseted dhs pos dataset with the same chromosome and which emcompases the position of the SNP in that row and extract the index number from that. Then put the indices into the empty mpra columns. You need the three columns for varaints with multiple indicies
#num_index<-vector()
for(i in 1:nrow(mpra)){
  #mpra[i,]$peak_index<-subset(dhs.pos, seqname==mpra[i,]$chr & start<=mpra[i,]$snp_end & end >=mpra[i,]$snp_end)$index[1]  
  indices<-subset(dhs.pos, seqname==mpra[i,]$chr & start<=mpra[i,]$snp_end & end >=mpra[i,]$snp_end)$index
  mpra[i,c("peak_index1","peak_index2","peak_index3")]<-c(indices, rep(NA, 3-length(indices)))
  #num_index<-c(num_index,length(subset(dhs.pos, seqname==mpra[i,]$chr & start<=mpra[i,]$snp_end & end >=mpra[i,]$snp_end)$index))
}

# list out the cells which are relevant to this cell type (tcell related). This is different than the bcell list which is used with the GM12878 data. The large list of cell types can be found in sample.dat in the Biosample.name column. 
cells<-c( "hTH1", "hTH17", "hTH2","CD4", "CD4pos_N", "hTR", "Jurkat", "CD8")

# Second loop
# For each cell type,
for(c in cells){
  # get the numbers of the cell types in the sample data
  cell_col<-as.vector((subset(sample.dat, Biosample.name==c)$library.order)+3)
  # If the length of these columns are greater than 1, then use the max dhs values from the dhs data 
  if(length(cell_col)>1){
    dhs.dat$max_dhs<-apply(dhs.dat[,cell_col], 1, max)
  }else{
  # Or just make the only cell type columns's dhs data the max dhs
    dhs.dat$max_dhs<-dhs.dat[,cell_col]
  }
  # Create a peak index of all the values that are greater than 0 
  peak_index<-subset(dhs.dat, max_dhs>0)$index

 # mpra$dhs_peak<-ifelse(mpra$peak_index%in%peak_index, 1, 0)
  # Make a column in the mpra that that contains a 1 if that that variant has a peak and a 0 if it does not
  mpra$dhs_peak<-ifelse(mpra$peak_index1%in%peak_index | mpra$peak_index2%in%peak_index| mpra$peak_index3%in%peak_index, 1, 0)
  # Make the name of that column specific to the specific cell type tested
  names(mpra)<-gsub("dhs_peak", paste0("dhs_", c), names(mpra))

}
# Create a column of dhs peaks that merges all of the DHS columns 
mpra$dhs_Tcell_merged<-apply(mpra[,which(grepl("dhs_",names(mpra)))],1, max)
# Create a column of dhs peaks from the original dhs pos column. How is this different from the column we just made?
mpra$dhs_all<-ifelse(mpra$peak_index1%in%dhs.pos$index | mpra$peak_index2%in%dhs.pos$index | mpra$peak_index3%in%dhs.pos$index, 1, 0)
# eliminate the peak columns
mpra<-subset(mpra, select=-c( peak_index1, peak_index2, peak_index3))
# Remove the objects you do not need anymore 
rm(mpra.se)
rm(dhs.pos.se)
```

```{r}
nrow(mpra)
```

deltaSVM
```{r, eval=FALSE}
#Merge in deltaSVM
# Load the Jurkat deltaSVM data
dat.deltasvm<-read.delim(paste0("/nfs/jray/screens/Mouri_et_al_MPRA/delta_svm/mpra_snps_E2_Jurkat_deltaSVM.txt"), header=F, stringsAsFactors = F, sep="\t")
# Label the names of the Jurkat deltaSVM data
names(dat.deltasvm)<-c("SNP", "delta_svm_Jurkat")
# merge the jurkat deltaSVM data with mpra
mpra<-merge(mpra, dat.deltasvm, by="SNP", all.x=T, all.y=F)
# Load the CD4 deltaSVM data
dat.deltasvm<-read.delim(paste0("/nfs/jray/screens/Mouri_et_al_MPRA/delta_svm/mpra_snps_E2_naive_CD4_deltaSVM.txt"), header=F, stringsAsFactors = F, sep="\t")
# Rename the columns
names(dat.deltasvm)<-c("SNP", "delta_svm_nCD4")
# merge with MPRA
mpra<-merge(mpra, dat.deltasvm, by="SNP", all.x=T, all.y=F)
```

```{r}
nrow(mpra)
```

ASC 
```{r, eval=FALSE}
#Merge in ASC from Calderon
#Read in allelic skew data from Calderon et al. (PMID: 31570894), Supplementary Table 10, "significant_ASCs" tab
#asc_dir<-"/Users/michaelguo/Dropbox/JR_MPRA/asc/"
# Import the spreadsheet into R
asc<-data.frame(read_excel("/nfs/jray/screens/Mouri_et_al_MPRA/asc/41588_2019_505_MOESM3_ESM.xlsx", sheet=10), stringsAsFactors = F)
# remove cell tpyes with have -S in them (aka stimulated cell types)
asc<-asc[!grepl("\\-S", asc$cell_type),]
# Extracting the chromosome from the ID and making it a column
asc$chr<-str_split_fixed(asc$het_id, "\\_", 5)[,2]
# # Extracting the position from the ID and making it a column
asc$pos<-as.numeric(str_split_fixed(asc$het_id, "\\_", 5)[,3])
# Empty column for this data in the MPRA (full of zeros)
mpra$asc<-0
# For each row in the mpra data, if the asc has a matching chromosome and position then the mpra row gets a 1
for(i in 1:nrow(mpra)){
  if(nrow(subset(asc, chr==mpra[i,]$chr & pos==mpra[i,]$snp_end))>0){
    mpra[i,]$asc<-1
  }
}
```

```{r}
nrow(mpra)
```

ATAC-QTL and eQTL (This chunk adds two extra rows wtf is going on)
```{r, eval=FALSE}
# Read in Gate ATAC-QTL data
# Read in Gate et al ATAC-QTL data (PMID: 29988122), Supplemental Table 6
# qtl_dir<-"/Users/michaelguo/Dropbox/JR_MPRA/qtl/"
# Read in supplementary table
qtl<-data.frame(read_excel("/nfs/jray/screens/Mouri_et_al_MPRA/qtl/41588_2018_156_MOESM8_ESM.xlsx", sheet=2), stringsAsFactors = F)
# Create a chromosome column from the SNP column
qtl$chr<-str_split_fixed(qtl$SNP, "\\:", 2)[,1]
# Create a position column from the SNP column
qtl$snp_end<-str_split_fixed(str_split_fixed(qtl$SNP, "\\_", 2)[,1], "\\:", 2)[,2]
# Subsetting to only the information you want to put in  the mpra
qtl<-subset(qtl, select=c(chr, snp_end, beta, p.value))
# Changing the column names to the column names you want to put in the mpra
names(qtl)[3:4]<-c("atac_qtl_beta", "atac_qtl_pval")
# Merge the atac qtl data with the mpra
mpra<-merge(mpra, qtl, by=c("chr", "snp_end"), all.x=T, all.y=F)

# Read in Gate eQTL data
# Read in the second supplementary table. This one is eqtls not atac-qtls. 
qtl<-data.frame(read_excel("/nfs/jray/screens/Mouri_et_al_MPRA/qtl/41588_2018_156_MOESM10_ESM.xlsx", sheet=2), stringsAsFactors = F)
# Create a chromosome column from the SNP column
qtl$chr<-str_split_fixed(qtl$SNP, "\\:", 2)[,1]
# Create a position column from the SNP column
qtl$snp_end<-str_split_fixed(str_split_fixed(qtl$SNP, "\\_", 2)[,1], "\\:", 2)[,2]
#qtl$chr_pos<-paste(qtl$chr, qtl$pos, sep="_")
# Subsetting to only the information you want to put in  the mpra
qtl<-subset(qtl, select=c(chr, snp_end, beta, p.value, gene))
# Changing the column names to the column names you want to put in the mpra
names(qtl)[3:5]<-c("eqtl_beta", "eqtl_pval", "eqtl_gene")
# Merge the eqtl data with the mpra
mpra<-merge(mpra, qtl, by=c("chr", "snp_end"), all.x=T, all.y=F)
# Eliminate the duplicated SNPs created by this merge
mpra <- subset(mpra, duplicated(SNP)!=TRUE)
```

```{r}
nrow(mpra)
```

MotifbreakR
```{r motifbreakr, eval=FALSE}
#Read in motifbreakr data
dat.motifbreakr<-read.delim(paste0("/nfs/jray/screens/Mouri_et_al_MPRA/mhguo1-T_cell_MPRA-5c36361/tf/motifbreakr/mpra.motifbreakr.method_log.p1e5.results.txt"),header=T, stringsAsFactors = F,sep="\t")
# Create a SNP column from the SNP_id column in the motifbreakr data
dat.motifbreakr$SNP<-gsub("chr", "", dat.motifbreakr$SNP_id)
# Subset the motifbreakr to only the SNP and the gene
dat.motifbreakr<-subset(dat.motifbreakr,  select=c(SNP,  geneSymbol))
# Create an empty column in the mpra for the motifbreakr data
mpra$tf_motifbreakr<-NA
# For each row in mpra, if the row's SNP is in the MPRA data set, subset the motifbreakr data to only that SNP and paste the genes as characters in the column with a comma in between and make that the tf_motifbreakr column in the mpra
for(i in 1:nrow(mpra)){
  if(mpra[i,]$SNP%in%dat.motifbreakr$SNP){
    mpra[i,]$tf_motifbreakr<-paste(subset(dat.motifbreakr, SNP==mpra[i,]$SNP)$geneSymbol, collapse=",")
  }
}
```

```{r}
nrow(mpra)
```

Nearest TSS
```{r genes from online, eval=FALSE}
#Add in genes
# Load the dataset. This takes a while. This dataset is in hg19
dat.genes<-fread("ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_19/gencode.v19.annotation.gtf.gz", header=F,data.table=F, skip=5)
# Subset the data to only genes and pick the correct columns you need
dat.genes<-subset(dat.genes, V3=="gene", select=c(V1, V4, V5, V7,V9))
# Give names to the columns
names(dat.genes)<-c("chr", "start", "end","strand", "annotation")
# Select only protein coding genes from the annotation
dat.genes<-dat.genes[grepl("protein_coding", dat.genes$annotation),]
# extract the gene name from the annotation
dat.genes$gene<-str_split_fixed(dat.genes$annotation, "\\;",10)[,5]
# eliminate the words gene_name from the gene column
dat.genes$gene<-gsub("gene_name ", "",dat.genes$gene)
# Eliminate the quotes from the gene name
dat.genes$gene<-gsub('"', "", dat.genes$gene, fixed=TRUE)
# Eliminate the spaces from the gene name
dat.genes$gene<-gsub(' ', "", dat.genes$gene, fixed=TRUE)
# Create a new column of tss which is just the start of the gene
dat.genes$tss<-dat.genes$start
# For every row in this ebi dataset, if the gene is on the negative strand, then the value for the tss column is actually the value for the end column because that would be the tss
for(i in 1:nrow(dat.genes)){
  if(dat.genes[i,]$strand=="-"){
    dat.genes[i,]$tss<-dat.genes[i,]$end
  }
}
# Eliminate the annotation and strand columns
dat.genes<-subset(dat.genes, select=-c(annotation, strand))

#Add nearest tss
# Create an empty column in the mora for this data 
mpra$tss<-NA
# For each row in mpra, subset the tss data so that it is the same chromosome as the row, then create a new column which is the absolute value of the difference of all the tss of that chromosome and and the variant in that row. Then order the data by the distance and pick the row with the shortest ditance to put in that row's tss column because that is the gene with the closest tss 
for(i in 1:nrow(mpra)){
   temp<-subset(dat.genes, chr==mpra[i,]$chr)
   temp$distance<-abs(temp$tss-mpra[i,]$snp_end)
   temp<-temp[order(temp$distance),]
  mpra[i,]$tss<-temp[1,]$gene
 }

```
```{r}
nrow(mpra)
```

This adds the rsID to the data set. 
```{r, eval=FALSE}
# read in one of the mouri et al. supplementary tables
rsid.dat<-read_excel("/nfs/jray/screens/Mouri_et_al_MPRA/Mouri_et_al_replication/MPRA_merge_creation/41588_2022_1056_MOESM4_ESM.xlsx")
# renaming a column to merge it with the mpra data
rsid.dat$SNP <- rsid.dat$ld_snp 
# pick just the SNP and rsid column
rsid.dat<-unique(subset(rsid.dat, select=c(SNP, rsid)))
# merge the rsid data with the mpra
mpra<-merge(mpra, rsid.dat, by="SNP", all.x=T, all.y=F)

missing_rsid_snps <- subset(mpra, is.na(rsid)==TRUE)
rsid_in_data <- missing_rsid_snps$ld_snp%in%rsid.dat$ld_snp
# missing_rsid_snps$ld_snp
```

```{r}
nrow(mpra)
```

Creating classifying mpra varaints as emVars, pCREs and no activity variants. This is the unstimulated jurkat filter from Mouri et al. 2022
```{r, eval=FALSE}
if(filter=="mouri"){
#  
# Manually temporarily set these FDRs and adjusted p-values that are NA to -1 for easier TRUE/FALSE testing. This is fine because p-values will never be -1.
mpra$Skew.logFDR[is.na(mpra$Skew.logFDR)]<- (-1)
mpra$A.logPadj_BH[is.na(mpra$A.logPadj_BH)]<- (-1)
mpra$B.logPadj_BH[is.na(mpra$B.logPadj_BH)]<- (-1)

#For each variant, assign whether it has no activity (nEnhancer_nSkew), is a pCRE only (Enhancer_nSkew) or emVAR (Enhancer_Skew). This is the original filter. The primary tcell filter will be assigned later.
# First assign every variant the no activity marker
mpra$mpra_sig<-"nEnhancer_nSkew"
# This filter calculates the variant category filter. For every row in the mpra data, if the skew FDR and FC BH p-value for allele A or B is greater than 1, then assign the variant as an emvar. If the skew FDr is less than 1 and the FC BH p-value for allele A or B is greater than 1, then assign the variant as a pcre. 
for(i in 1:nrow(mpra)){
  if(mpra[i,]$Skew.logFDR>1 & (mpra[i,]$A.logPadj_BH>1 |  mpra[i,]$B.logPadj_BH>1) ){
    mpra[i,]$mpra_sig<-"Enhancer_Skew"
  }else if(mpra[i,]$Skew.logFDR<=1 & (mpra[i,]$A.logPadj_BH>1 |  mpra[i,]$B.logPadj_BH>1) ){
    mpra[i,]$mpra_sig<-"Enhancer_nSkew"
  }
}

#Set those -1 FDRs and p-values back to NA
mpra$Skew.logFDR[mpra$Skew.logFDR<0]<- NA
mpra$A.logPadj_BH[mpra$A.logPadj_BH<0]<- NA
mpra$B.logPadj_BH[mpra$B.logPadj_BH<0]<- NA
#
}
```

This is John's code to check that the code selects variants that have a log2FC of enhancer activity > 1 for at least one allele per element.  It will create the mpra_sig column. There already is a column called mpra_sig and this should replace it. This uses the absolute value filter. If this filter is bad then I will have to rerun the correct code and all of the code after it. 
```{r, eval=FALSE, warning=FALSE}
if(filter=="tcell"){
#
# Create absolute value columns 
mpra$abs.A.log2FC <- abs(mpra$A.log2FC)
mpra$abs.B.log2FC <- abs(mpra$B.log2FC)
# Reset the mpra_sig column
mpra$mpra_sig<- NA
# For every row in the mpra, if the absolute value of the fold change is above 1 and the BH p-vlaue is above 2 for either allele and the LogSkew FDR is above 1, then the variant is labeled as an emVar (Enhancer_Skew). Else, if the absolute value of the fold change is above 1 and the BH p-vlaue is above 2 for either allele, then it is labeled as a pCRE (Enhancer_nSkew). If it is niether of those it is a no activity variant (nEnhancer_nSkew). 
for(i in 1:nrow(mpra)){
  mpra[i,]$mpra_sig<-ifelse(((mpra[i,]$abs.A.log2FC>1 &mpra[i,]$A.logPadj_BF>2) | (mpra[i,]$abs.B.log2FC>1 & mpra[i,]$B.logPadj_BF>2)) & mpra[i,]$Skew.logFDR>=1, "Enhancer_Skew", ifelse((mpra[i,]$abs.A.log2FC>1 &mpra[i,]$A.logPadj_BF>2) | (mpra[i,]$abs.B.log2FC>1 & mpra[i,]$B.logPadj_BF>2), "Enhancer_nSkew", "nEnhancer_nSkew"))
}
# 
}
```

```{r}
nrow(mpra)
```

```{r}
z <- read.table("/nfs/jray/screens/ALL_MPRAs/Ho_et_al_analysis/Downstream_Analysis/20240310_mpra_analysis/tcell/glm/data/20240310_tcell_glm_test", sep="\t", header=T)
```

This will remove the bad SNPS and MHC/HLA
```{r, eval=FALSE}
nrow(mpra)
# Eliminate this SNP because it had issue in the sequencing library. 
bad_snp <- subset(mpra, rsid == "rs5855282")
bad_snp$SNP
mpra <- mpra[mpra$SNP != "3:189662658:GT:G",]
nrow(mpra)
missing_rsid_snps <- subset(mpra, is.na(rsid)==TRUE)
mpra$bad_snps <- mpra$ld_snp%in%missing_rsid_snps$ld_snp

# bad_snps2 <- missing_rsid_snps$SNP
mpra <- subset(mpra, bad_snps == FALSE)
mpra <- subset(mpra, select=-c(bad_snps))
nrow(mpra)
```

```{r}
nrow(mpra)
```

Final checks before writing the mpra merge file 
```{r}
# names(mpra)
subset_mpra <- subset(mpra, mpra_sig == "Enhancer_Skew")
nrow(subset_mpra)
# emVars (tcell glm) (568 with extra 40 variants)
subset_mpra2 <- subset(mpra, mpra_sig == "Enhancer_nSkew")
nrow(subset_mpra2)
# 1128 pCRES (tcell glm) (1144 with extra 40 variants)
subset_mpra3 <- subset(mpra, mpra_sig == "nEnhancer_nSkew")
nrow(subset_mpra3)
# 16604 no activity (tcell glm) (16607 with extra 40 variants)
```


Write table 
```{r,eval=FALSE}
write.table(mpra, paste0(mpra.dir,"/data/",mpra.name), row.names=F, col.names=T, sep="\t", quote=F)
```

Bring in the table again
```{r}
mpra <- read.table(paste0(mpra.dir,"/data/",mpra.name), sep="\t", header=T)
```

Add top_pics and top_disease
```{r}
#For each MPRA variant, find the pics for all the diseases and make max_pics the maximum of all of those values
mpra.pics <- subset(mpra, select=c(MS_pics,RA_pics,UC_pics,T1D_pics,IBD_pics,Crohns_pics,Psoriasis_pics))

# Make the NAs for pics -1
mpra.pics$MS_pics[is.na(mpra.pics$MS_pics)]<- (-1)
mpra.pics$RA_pics[is.na(mpra.pics$RA_pics)]<- (-1)
mpra.pics$UC_pics[is.na(mpra.pics$UC_pics)]<- (-1)
mpra.pics$IBD_pics[is.na(mpra.pics$IBD_pics)]<- (-1)
mpra.pics$T1D_pics[is.na(mpra.pics$T1D_pics)]<- (-1)
mpra.pics$Crohns_pics[is.na(mpra.pics$Crohns_pics)]<- (-1)
mpra.pics$Psoriasis_pics[is.na(mpra.pics$Psoriasis_pics)]<- (-1)
colnames(mpra.pics)
mpra.pics$win_pics <- colnames(mpra.pics)[apply(mpra.pics,1,which.max)] # Make a column that is called win_pics which will give me the name of the pics column with the best 

mpra.pics$max_pics <- NA # This creates a column full of NAs for the loop to fill

for(i in 1:nrow(mpra.pics)){
  if(mpra.pics[i,]$win_pics=="MS_pics"){
    mpra.pics[i,]$max_pics<-mpra.pics[i,]$MS_pics
  } # if win_pics says MS_pics then the max_pics is MS_pics
  if(mpra.pics[i,]$win_pics=="RA_pics"){
    mpra.pics[i,]$max_pics<-mpra.pics[i,]$RA_pics
  } # if win_pics says RA_pics then the max_pics is RA_pics
  if(mpra.pics[i,]$win_pics=="UC_pics"){
    mpra.pics[i,]$max_pics<-mpra.pics[i,]$UC_pics
  } # if win_pics says UC_pics then the max_pics is UC_pics
  if(mpra.pics[i,]$win_pics=="IBD_pics"){
    mpra.pics[i,]$max_pics<-mpra.pics[i,]$IBD_pics
  } # if win_pics says IBD_pics then the max_pics is IBD_pics
  if(mpra.pics[i,]$win_pics=="Crohns_pics"){
    mpra.pics[i,]$max_pics<-mpra.pics[i,]$Crohns_pics
  } # if win_pics says Crohns_pics then the max_pics is Crohns_pics
  if(mpra.pics[i,]$win_pics=="Psoriasis_pics"){
    mpra.pics[i,]$max_pics<-mpra.pics[i,]$Psoriasis_pics
  } # if win_pics says Psoriasis_pics then the max_pics is Psoriasis_pics
   if(mpra.pics[i,]$win_pics=="T1D_pics"){
    mpra.pics[i,]$max_pics<-mpra.pics[i,]$T1D_pics
  } # if win_pics says T1D_pics then the max_pics is T1D_pics
}

# Make the MPRA pics columns which are NA return to NA
mpra.pics$MS_pics[mpra.pics$MS_pics<0]<- NA
mpra.pics$RA_pics[mpra.pics$RA_pics<0]<- NA
mpra.pics$UC_pics[mpra.pics$UC_pics<0]<- NA
mpra.pics$IBD_pics[mpra.pics$IBD_pics<0]<- NA
mpra.pics$T1D_pics[mpra.pics$T1D_pics<0]<- NA
mpra.pics$Crohns_pics[mpra.pics$Crohns_pics<0]<- NA
mpra.pics$Psoriasis_pics[mpra.pics$Psoriasis_pics<0]<- NA

```

Putting the max_pics and win_pics back into mpra
```{r}
mpra.pics$top_disease <- str_extract(mpra.pics$win_pics, "[^_]+")
mpra$top_disease <- mpra.pics$top_disease 
mpra$top_pics <- mpra.pics$max_pics
```

```{r}
nrow(mpra)
```

Write table 
```{r,eval=FALSE}
write.table(mpra, (paste0(mpra.dir,"/data/",mpra.name)), row.names=F, col.names=T, sep="\t", quote=F)
```

Bring in the table again
```{r}
mpra <- read.table(paste0(mpra.dir,"/data/",mpra.name), sep="\t", header=T)
```

MPRA merge checks for DHS
```{r}
original_mpra <-read.delim("/nfs/jray/screens/ALL_MPRAs/Ho_et_al_analysis/Downstream_Analysis/2020_mouri_public_data_replication/data/mpra_data_merge.txt", header=T, stringsAsFactors = F, sep="\t")
original_mpra_dhs <- subset(original_mpra, select=c(SNP,dhs_CD4,dhs_Jurkat,dhs_CD8,dhs_Tcell_merged,dhs_all))
mpratest <- subset(mpra, select=c(SNP,dhs_CD4,dhs_Jurkat,dhs_CD8,dhs_Tcell_merged,dhs_all))
dhs_merged <- merge(mpratest, original_mpra_dhs, by=c("SNP"))
identical(dhs_merged[["dhs_CD4.x"]],dhs_merged[["dhs_CD4.y"]])
# FALSE! 

dhs_merged$results = ifelse(dhs_merged$dhs_CD4.x > dhs_merged$dhs_CD4.y, 'Column1', 
                      ifelse(dhs_merged$dhs_CD4.x < dhs_merged$dhs_CD4.y, 'Column2', 'None'))

# There should only be "None" in the column we just created

class(dhs_merged$dhs_CD4.x)
class(dhs_merged$dhs_CD4.y)

# It is okay if one is an integer and one is a numeric 
```


Find the extra variants in the MPRA which are in the 
```{r}
# Collection of SNPs in MPRA
set1 <- paste(mpra$SNP)
# Selection of SNPs in the merged MPRA
set2 <- paste(mpra_out$SNP)
# Create a venn diagram of the two SNP list to see which SNPs don't overlap
library(ggvenn)
B <-list("mpra merge"=set1, ".out file"=set2)
names(B) <- c("mpra merge",".out file")
venn_diagram_plot <- ggvenn(B, set_name_size = 5, fill_color = c("springgreen2","forestgreen"),show_elements = TRUE, label_sep = "\n") + 
  ggtitle("mpra merge vs. .out file") +
  theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"))
venn_diagram_plot
mpra[9810,]
mpra$SNP%in%mpra_out$SNP
xtra <- mpra$SNP[duplicated(mpra$SNP)]
extra_snps <- nrow(subset(mpra, is.na(SNP)==TRUE))
```

